---
title: "Writing (good) functions in R"
author: "Adam Black"
date: "2/11/2020"
output:   
  ioslides_presentation:
    widescreen: TRUE
---

<!-- HTML to enable scrolling in slides -->
<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
</style>


```{r setup, include=FALSE, message=F}
knitr::opts_chunk$set(echo = T, eval = F)
library(tidyverse)
```

## Disclaimer

This presentation is copied verbatim from an old datacamp course on writing functions.

## Parts of a function

```{r}
name <- function(arg1, arg2) {
  # body
}
```

**Exercise 1:** Type "fun" and push tab in RStudio to insert a function template.

## Arguments can be specified by name or position

```{r, eval = T}
ratio <- function(x, y) {
  x/y
}

# match arguments by position
ratio(3, 4)
# match arguments by name
ratio(y = 4, x = 3)
```

The last evaluate expression in the function is what the function returns.


## Environments

Names are first looked up in a function's arguments. If not found they will be looked up in the environment where the function was defined (often the global environment).

```{r, eval = T}
x <- 100
ratio <- function(x, y) {
  x/y
}

ratio(3, 4)

# Normally we want to avoid code like this 
z <- 100
ratio <- function(x, y) {
  z/y
}

ratio(3, 4)

```

The last evaluate expression in the function is what the function returns.

## Why write a function?

- Make intent of code more clear to the reader
- Reduces mistakes from copy and pasting code
- Less duplicated code is easier to change

## Why write a function?

- What is the intent of the code below?
- How was it written?
- Can you spot the error?

```{r, eval = F}
df$a <- (df$a - min(df$a, na.rm = TRUE)) /
    (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))

df$b <- (df$b - min(df$b, na.rm = TRUE)) /
    (max(df$a, na.rm = TRUE) - min(df$b, na.rm = TRUE))

df$c <- (df$c - min(df$c, na.rm = TRUE)) /
    (max(df$c, na.rm = TRUE) - min(df$c, na.rm = TRUE))

df$d <- (df$d - min(df$d, na.rm = TRUE)) /
    (max(df$d, na.rm = TRUE) - min(df$d, na.rm = TRUE))

```

## When should you write a function?

If you have copied and pasted code twice, it's time to write a function.

```{r, eval = F}
df$a <- rescale01(df$a)
df$b <- rescale01(df$b)
df$c <- rescale01(df$c)
df$d <- rescale01(df$d)
```

## How to write a function?

**Don't start with the function template!**

Instead follow a 5 step process.

- Start with a simple problem
- Get a working snippet of code
- Rewrite to use temporary variables
- Rewrite for clarity
- Finally turn the snippet into a function

## 1. Start with a simple problem

Your problem should be concrete and you should know the answer.

```{r, eval = T}
df <- data.frame(
  a = 1:11,
  b = rnorm(11),
  c = rnorm(11),
  d = rnorm(11)
)

# Rescale thea column in df to the interval [0,1]

# Output should be
 # [1] 0.0 0.1 0.2 0.3 0.4 0.5 0.6 0.7 0.8 0.9 1.0
```



## 2. Get a working snippet of code that solves the problem

Examine the working code snippet to determine what the function inputs should be

```{r}

df$a <- (df$a - min(df$a, na.rm = TRUE)) /
    (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))

```



## 3. Rewrite the snippet using temporary variables

Examine the working code snippet to determine what the function inputs should be

```{r}
x <- df$a

(x - min(x, na.rm = TRUE)) /
    (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))

```


## 4. Rewrite for clarity

```{r}
x <- df$a

# extract out calculation of the range
rng <- range(x, na.rm = T)

(x - rng[1]) / (rng[2] - rng[1])

```

## 5. Finally, turn your working snippet into a function

Make sure all teporary variables become arguments.
Test the function on your simple problem as well as other similar problems.

```{r, eval = T}
x <- df$a

rescale01 <- function(x) {
  rng <- range(x, na.rm = T)
  (x - rng[1]) / (rng[2] - rng[1])
}

# Test the function
rescale01(x)
```


## Use your function just like any other R function

```{r, eval = T}
df %>% 
  mutate_all(rescale01)
```




## How to write a good function?

- A function should solve a problem **correctly**
- A function should be understanable to humans
- Correct + understandable = obviously correct


## What does this code do?
```{r, eval = F}
baz <- foo(bar, bla)
```

## What does this code do?
```{r}
df2 <- arrange(df, time)
```


## Naming principles

- Pick a consistent style for long names
```{r}
# Good
col_mins()
row_maxes()

# Bad
newData <- c(old.data, todays_log)
```


- Do not override existing variables or functions
```{r}
# Bad
T <- FALSE
c <- 10
mean <- function(x) sum(x)
```



## Function names

- Should generally be verbs 

```{r}
# Good
impute_missing()

# Bad
imputed()
```

- Should be descriptive
```{r}
# Good
collapse_years()

# Bad
f()
my_awesome_function()
```

## Argument names

- Should generally be nouns
- Use very common short names when appropriate:
  * x, y, z: vectors
  * df: a dataframe
  * i, j: numeric indicies (typically rows and columns)
  * n: length or number of rows
  * p: number of columns


## Argument order

Arguments generally fall into two categories

- **data arguments** supply data to compute on
- **detail arguments** control the details of the computations

```{r}
mean(x, trim = 0, na.rm = FALSE, ...)

t.test(x, y = NULL,
       alternative = "two.sided",
       mu = 0, paired = FALSE, var.equal = FALSE,
       conf.level = 0.95, ...)
```

Exercise: What are the data and detail arguments in the function calls above?

## Argument order

- Data arguments should come before detail arguments
- Detail arguments should have sensible default values



## How to write a function?
 
- Start with a simple problem
- Get a working snippet of code
- Rewrite to use temporary variables
- Rewrite for clarity
- Finally turn the snippet into a function

## What makes a good function?
- Use good names for functions and arguments
- Use an intuitive argument order and reasonable default values
- Make it clear what the function returns
- Use good code style inside the body of the function

https://style.tidyverse.org/


## Writing functions with dplyr code

dplyr uses non-standard evaluation to make it easy to use for analysts. However this can causes problems when using dplyr inside functions. The solution to this problem is *tidyeval* which is a topic for another day.

If you find yourself needing to write functions with dplyr check out these links.

https://dplyr.tidyverse.org/articles/programming.html

https://edwinth.github.io/blog/dplyr-recipes/


## Time to practice!

The airquality dataset has daily readings of the following air quality values for May 1, 1973 to September 30, 1973.

Our problem is to replace missing values in each column with the median value in that column.


Use the 5 steps to write a function that imputes missing data with the median non-missing value in the data.

Use your function to impute all missing data in the built-in airquality dataset. Next run your function on the sleep dataset in the VIM package.




